---
title: "Visualizing Water: Sankey Diagrams (draft)"
description: |
  Part 2: custom data-driven flow charts using Highcharts.js and networkD3 (IWMI).
author:
  - name: Melanie BACOU
    email: mel@mbacou.com
    url: https://linkedin/in/mbacou
categories:
  - imwi
  - hydrology
date: "`r Sys.Date()`"
date_prefix: 2021-10-01
---

```{r setup, include=FALSE}

library(knitr)
library(jsonlite)
library(scales)
library(stringr)
library(data.table)
library(highcharter)

opts_chunk$set(res=220, pointsize=8)

# Default color palette
pal <- readRDS("../../_assets/pal_iwmi.rds")

```


This notebook is Part 2 of a code playground for hydrologic model visualization. In [Part 1](../2021-10-16-svg/) we focused on `D3.js`, here we test network and hierarchical visuals using higher-level graphic libraries [Highchart.js](https://www.highcharts.com/) and [networkD3](http://christophergandrud.github.io/networkD3/).

The 2 libraries are built on top of D3 but they come with sample configurations and R and Python clients, making it easier to build dynamic charts, esp. Sankey, treemaps, network and flow diagrams with significantly less code.

The approach here is inspired by:

1. TruthStudio for the Nature Conservancy's [static illustrations](https://www.truthstudio.com/portfolio_page/visualizing-water/) of the Colorado River basin's withdrawals and consumption.
2. [Curmi et al. 2013](https://www.researchgate.net/publication/257673216_Visualising_a_Stochastic_Model_of_Californian_Water_Resources_Using_Sankey_Diagrams) 
3. Lancaster University [Food-Energy Flows](https://github.com/csaladenes/food-energy-flows) diagram


```{r, out.width="50%", fig.width=3, fig.show="hold"}

c(
  "./fig/sankey_truthstudio_1.gif",
  "./fig/sankey_truthstudio_2.gif",  
  "./fig/sankey_curmi_2013.png",
  "./fig/sankey_lancaster.png"
) %>%
  include_graphics()

```


```{r}

# Load yearly dataset generated in Part 1
data <- fread("../2021-10-16-svg/data/data.csv")

```


# Sankey Diagram {#sankey}

A Sankey diagram is a type of directed network graph in which the width of the arrows (vertices) is proportional to the flow rate.

Here we use R library [`networkD3`](http://christophergandrud.github.io/networkD3/) to build the charts (chart configurations built in R can be exported to vanilla Javascript). First we define a default theme for Highcharts, which we'll reuse across visuals.

```{r}

hc_theme_iwmi <- function(
  hc,
  title = NULL,
  subtitle = NULL,
  label = NULL,
  x = NULL,
  y = NULL,
  exporting = TRUE,
  credits = FALSE,
  ...) {
  
  thm <- hc_theme(
    
    chart = list(
      backgroundColor = "transparent"
    ),
    # Don't use semantic colors
    colors = unname(pal),
    title = list(
      style = list(color="#333", fontSize="16px"),
      align = "left"
    ),
    subtitle = list(
      style = list(color='#777', fontSize="13px"),
      align = "left"
    ),
    legend = list(
      enabled = TRUE,
      itemStyle = list(color="#777"),
      verticalAlign = "top",
      align = "left",
      itemHoverStyle = list(color="#333")
    ),
    xAxis = list(
      title = list(enabled=FALSE),
      dateTimeLabelFormats = list(day='%e %b', week='%e %b %y', month='%b-%y', year='%Y'))
    ,
    yAxis = list(
      title = list(enabled=FALSE)
    ),
    tooltip = list(
      enabled = TRUE, shared = TRUE, split = FALSE,
      pointFormat = "{series.name}: <strong>{point.y:,.1f}</strong><br/>",
      xDateFormat = "%Y-%m-%d",
      dateTimeLabelFormats = "%Y-%m-%d",
      valueDecimals = 2
    ),
    plotOptions = list(
      series = list(
        opacity = .8,
        connectNulls = TRUE,
        marker = list(enabled=NA, radius=3, enabledThreshold=8),
        dataLabels = list(enabled=NA, style=list(fontSize="11px"))
      ),
      heatmap = list(
        marker = list(enabled=TRUE, lineWidth=6, lineColor="#fff"),
        dataLabels = list(enabled=TRUE, pointFormat="{point.value:,.0f}")
      )
    ),
    exporting = list(
      enabled = exporting,
      csv = list(dateFormat="%Y-%m-%d"),
      buttons = list(contextButton=list(
        symbolSize = 12,
        symbolFill = "#fff",
        symbolStroke = "#777",
        symbolStrokeWidth = 1.3,
        menuItems = c("printChart", "downloadPNG", "downloadSVG", "downloadCSV")))
    ),
    credits = list(
      enabled = credits,
      position = list(align="left"),
      href = "https://iwni.cgiar.org/",
      style = list(fontSize="11px")
    ),
    ...
  )
  
  p  = hc_add_theme(hc, thm) %>%
    hc_title(text = title) %>%
    hc_subtitle(text = subtitle)
  
  if(length(label)>0) p = p %>%
    hc_annotations(list(labels = list(
      list(text=label, useHTML=TRUE, shape="rect", point=list(x=x, y=y)))))
  
  return(p)
}

```

Then we reshape our sample Kenya dataset for 2017 into a list of vertices between 2 nodes (with fields `from`, `to`, `value` at minimum).

```{r}



```



# Steamgraphs


