---
title: Custom Highcharts (draft)
description: |
  Part 2: convert WA+ sheets to custom data-driven charts using Highcharts.js (IWMI).
author:
  - name: BACOU, Melanie
    email: mel@mbacou.com
    url: https://linkedin/in/mbacou
preview: resource-base.svg
categories:
  - draft
  - imwi
  - hydrology
date: "`r Sys.Date()`"
date_prefix: 2021-10-01
output:
  distill::distill_article:
    toc: true
    dev: svglite
    code_folding: true
    self_contained: false
---

```{r setup, include=FALSE}

library(knitr)
library(jsonlite)
library(scales)
library(stringr)
library(data.table)
library(highcharter)

opts_chunk$set(res=220, pointsize=8)

# Default color palette
pal <- readRDS("../../_assets/palette.rds")

```


This notebook is Part 2 of a code playground for hydrologic model visualization. In [Part 1](../2021-10-16-svg/) we focused on `D3.js`, here we build Javascript widgets using a higher-level graphic library `Highchart.js`.

`Highcharts` is built on top of D3 but comes with sample configurations and R and Python clients, making it easier to build highly dynamic charts, esp. Sankey, treemaps, flow diagrams, etc. with less code.

# Data Preparation

To validate the different approaches we use sample model output and datasets provided by **Naga Valpuri** for Mali (Niger river basin) and Kenya (Mara river  basin). 

```{r}

data <- list(
  ken = "~/Projects/2021-iwmi/data/ken/hydroloop_results/csv",
  mli = "~/Projects/2021-iwmi/data/mli/csv_km3"
) %>%
  lapply(list.files, pattern="*.csv", recursive=TRUE, full.names=TRUE) %>%
  lapply(data.table) %>%
  rbindlist(idcol="iso3", use.names=TRUE, fill=TRUE) %>%
  setnames("V1", "path")

data[, 
  file := basename(path)
][, `:=`(
  year = str_extract(file, "_[0-9]{4}") %>% str_sub(2,5) %>% as.integer(),
  month = str_extract(file, "[0-9]{4}_[0-9]{1,2}") %>% str_sub(6,7) %>% as.integer()
)] %>% setorder(iso3, year, month, na.last=TRUE)

data[iso3=="ken", .(file, year, month)][c(1:5, (.N-4):.N)] %>% 
  kable(caption="Kenya - List of output files (top and last 5)")
data[iso3=="mli", .(file, year, month)] %>% 
  kable(caption="Mali - List of output files")

```

In particular we have yearly model steps for Mali and monthly steps for Kenya that are further aggregated to seasonal and yearly time spans (by variable), as well as monthly time-series. In Mali only **Sheet 1** was produced (Resource Base).

A sample simulation output for Kenya for the year 2017.

```{r}

f <- "sheet1_2017.csv"
ken <- data[iso3=="ken" & file==f][1, fread(path)]
ken %>% 
  kable(caption=f)

```

This file lists `r nrow(ken)` output variables grouped into `CLASS` and `SUBCLASS`. Units are in **km³/year**. These numbers represent flow variables (water flowing into the basin, and water leaving the basin as both ground and surface waters, either stored, utilized, evaporated, infiltrated, or runoff).


# Sankey Diagram {#sankey}

A Sankey diagram is a type of directed network graph in which the width of the arrows (vertices) is proportional to the flow rate.

Here we use R library [`networkD3`](http://christophergandrud.github.io/networkD3/) to build the charts (chart configurations built in R can be exported to vanilla Javascript). First we define a default theme for Highcharts, which we'll reuse across visuals.

```{r}

hc_theme_iwmi <- function(
  hc,
  title = NULL,
  subtitle = NULL,
  label = NULL,
  x = NULL,
  y = NULL,
  exporting = TRUE,
  credits = FALSE,
  ...) {
  
  thm <- hc_theme(
    
    chart = list(
      backgroundColor = "transparent"
    ),
    # Don't use semantic colors
    colors = unname(pal),
    title = list(
      style = list(color="#333", fontSize="16px"),
      align = "left"
    ),
    subtitle = list(
      style = list(color='#777', fontSize="13px"),
      align = "left"
    ),
    legend = list(
      enabled = TRUE,
      itemStyle = list(color="#777"),
      verticalAlign = "top",
      align = "left",
      itemHoverStyle = list(color="#333")
    ),
    xAxis = list(
      title = list(enabled=FALSE),
      dateTimeLabelFormats = list(day='%e %b', week='%e %b %y', month='%b-%y', year='%Y'))
    ,
    yAxis = list(
      title = list(enabled=FALSE)
    ),
    tooltip = list(
      enabled = TRUE, shared = TRUE, split = FALSE,
      pointFormat = "{series.name}: <strong>{point.y:,.1f}</strong><br/>",
      xDateFormat = "%Y-%m-%d",
      dateTimeLabelFormats = "%Y-%m-%d",
      valueDecimals = 2
    ),
    plotOptions = list(
      series = list(
        opacity = .8,
        connectNulls = TRUE,
        marker = list(enabled=NA, radius=3, enabledThreshold=8),
        dataLabels = list(enabled=NA, style=list(fontSize="11px"))
      ),
      heatmap = list(
        marker = list(enabled=TRUE, lineWidth=6, lineColor="#fff"),
        dataLabels = list(enabled=TRUE, pointFormat="{point.value:,.0f}")
      )
    ),
    exporting = list(
      enabled = exporting,
      csv = list(dateFormat="%Y-%m-%d"),
      buttons = list(contextButton=list(
        symbolSize = 12,
        symbolFill = "#fff",
        symbolStroke = "#777",
        symbolStrokeWidth = 1.3,
        menuItems = c("printChart", "downloadPNG", "downloadSVG", "downloadCSV")))
    ),
    credits = list(
      enabled = credits,
      position = list(align="left"),
      href = "https://iwni.cgiar.org/",
      style = list(fontSize="11px")
    ),
    ...
  )
  
  p  = hc_add_theme(hc, thm) %>%
    hc_title(text = title) %>%
    hc_subtitle(text = subtitle)
  
  if(length(label)>0) p = p %>%
    hc_annotations(list(labels = list(
      list(text=label, useHTML=TRUE, shape="rect", point=list(x=x, y=y)))))
  
  return(p)
}

```

Then we reshape our sample Kenya dataset for 2017 into a list of vertices between 2 nodes (with fields `from`, `to`, `value` at minimum).

```{r}



```



# Treemaps {#treemap}

A barebone treemap to start with.


```{r, fig.cap="Sample Treemap", preview=TRUE}

# Add shaded colors
ken[, COLOR := abs(VALUE)] %>%
  data_to_hierarchical(c(CLASS, SUBCLASS, VARIABLE), COLOR, pal[c(2,4,3)]) %>%
  hchart(
    type = "treemap",
    layoutAlgorithm = "squarified",
    dataLabels = list(enabled=T, format="{point.name}"),
    tooltip = list(pointFormat="
      {point.parent}<br/>{point.name}<br/>Value: {point.value:,.2f}")
  ) %>%
  hc_theme_iwmi("Resource Base", "Kenya 2017") %>%
  hc_exporting(enabled=TRUE)

```


# Steamgraphs


