---
title: U.S. Veteran Statistics
subtitle: Subtitle comes here
description: |
  Explore 10-year trends in U.S. military casulaties.
date: 2022-12-01
date-modified: last-modified
draft: false
freeze: false
categories:
  - risk modeling
abstract: |
  This is is much longer abstract than description.
---


https://www.arlingtoncemetery.mil/Developers/Burial-Record-Public-Service-Methods

https://wspublic.eiss.army.mil/v1/IssRetrieveServices.svc/search?q=DOD=_:_:2022,DOB=_:_:1981&sortColumn=PrimaryLastName,PrimaryFirstName&sortOrder=asc

```{r}

library(ggplot2)


```

```{r}

plot(1:8, 2:9, t="o", col=1:8, cex=2)
grid()
lattice::show.settings()

```

```{r cover}

ggplot(airquality, aes(x=Temp, fill=factor(Day))) + 
  geom_histogram(color="white") +
    ggtitle("Temperature Mean of Min -- ERA5 vs. GHCN (19 WS)") +
  theme(
    legend.position="top", legend.justification=0, 
    axis.text.x=element_text(angle=-90)
  )

```

::: {.callout-note}
Note that there are five types of callouts, including:
`note`, `warning`, `important`, `tip`, and `caution`.
:::

::: {.callout-tip}
## Tip with Title

This is an example of a callout with a title.
:::

::: {.callout-caution collapse="true"}
## Expand To Learn About Collapse

This is an example of a 'folded' caution callout that can be expanded by the user. You can use `collapse="true"` to collapse it by default or `collapse="false"` to make a collapsible callout that is expanded by default.
:::

::: {.callout-tip title="Tip with Title"}
This is a callout with a title.
:::
